Discussion Draft

### 2020-04-06 doc/Discussion.rmd
wraping up draft

We performed hypothesis testing of models from the literature and adapted here fit the main results from this study. We choose to examine three modifier models and two functional and cell physiology models (table X). It is a bit challenging to test all of these models together since they were built with difference parameters and built to describe different patterns of recombination variation. The modifier models were built to explain heterochiasmy variation between sexes and the COM functional model was built to describe the interference pattern and difference in the recombination landscape between the sexes. In cases where the models can't be extended to predicting results, we list as 'NA'.

@@

The direction of heterochiasmy under the **two locus modfier model** is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) that the driver acts (see Table 1 of @brandvain2012scrambling).

**connecting reverse HetC to faster male evolution**

**strict SAC drives directional selection**,  
**genetic variants which lead to 'unstable chromosome structure will be selected agaisnt**  
**sex differences in SAC strength are connected to conserved differences in gametogensis**

Under the **spindle based** model, faster male evolution of the genome wide recombination rate, (and instances of reversed heterochiasmy) is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle before starting the irreversable metaphase-to-anaphase transition (cite).  A stable chromosome structure requires, i) kinetochores attched to opposite poles of the cells and ii) at least one crossover to create tension across a propotion of sister cohesion (cite). Unstable bivalents on the MI spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If unstable bivalents fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, Sub cite). The different the detection of achiasmated bivalents (), the rates of annuploidy in mature gametes, and stronger infertility phenotypes in mutants of key SAC genes (cite).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite).
These results // Supporting this stage/cell physiology based mechanism  influencing broad sexual dimorphic phenotypes in recombination landscape. (leading to distinct evolutionary trajectories) **directional selection** in males and relaxed selection in females.

@@

Our results of greater between cell variation in females, these results are supported by human and mouse [@lynn2002, @gruhn2013]. While the **modifier models** parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the **spindle based** selection model, the less efficient SAC in oocytes leads to relaxed selection on the chromosome structure due to meiotic progression even with with unstable bivalents. This will result in increased variance in chromosome structure, and crossover number, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007). This is thought to reduce the triggering signal for the SAC.  This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite)).

<stabilization -- bi-oriented, robust SAC response, SAC acts to correctly time metaphase to anaphase transition -- only after correct alignment and biorientation of bivalents in the metaphase cell, -- spindle dynamics much different due to asymetrical diviation Mogessie et al 2018, Mogessie Schuh --actin and MT are apart of oocyte spindle>

<Bennabi et al 2016 --spindle assembly inoocytes, sun and kim 2011 -- SAC and regulators, Kolano et al 2012 --error prone oocytes, Manil-Segalen -- forces through MT breaking chromsomes, Mogessie et al 2018 -- meiotic spindle (big)review>

@@

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The **two locus modifier** and **SACE** model predicts sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The two mechanical models both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differening strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

Under the **spindle based model** we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (cite). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate (cite). Different timing of cell cycle between oogenesis and spermatogenesis may impose the the different selective pressures on sister cohesion. Faster spermatogenesis may select for synchronization of the separation homologs. While in oogenesis, the slower cell cycle and mulitple arrest stages may require chromosome structures with greater stability on the MI spindle.

@@

Our results from musculus^MSM^ and musculus^PWD^ show that the SC length and chromatin compaction is uncoupled from the direction of heterochiasmy. These results slightly depart from predictions which nominate chromatin organization (axis length) as the primary driver of heterochiasmy [@petkov2007].  While the relationship may hold for more heterochiasmy patterns, but these results indicate that rapid male specific  evolution in gwRR, is through a different trait than chromatin compaction).

<The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002, @gruhn2013], But there's generally a lack of data for other systems see [@cahoonLibuda2019]).>

Only the COM model predicts sexual dimorphism in chromatin structure will be longer in oocytes due to greater cell volume. This prediction / model might fit broader pattern such as in At where pollen is the larger cell and has longer axis length [@cahoonLibuda2019].

@@

Interference is a fundamental aspect genome-wide recombination rate the, chromatin compaction and the resulting recombination landscape.

Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength.

This pattern has empirical support from the most species [@ottoPaysuer2019] and fits well with  the fundamental relationship between the SC area or axis length, the physical upper limit for the number of crossovers. (**although it is not particulary strong in our results**).

**interference strength evolve**
**other examples of interferene strength evolution, tricky due to chromosome specific effects and complications from large evolutionary distance (change in underlying karyotype**

Our results find that interference strength has evolved in males from 3 strains from our panel of house mice.

Examples of evolution of interference strength are rare and complicated by changes in the underlying karyotype structure (Segura et al 2014).

The general patern of the relationship
 -- most findings / the general pattern of interference evolution is that it has a negative correlation to genome wide recombination rate. (Review in otto payseur), X examples -- correlation of X across x data.
--This fit logic of XX -- and also can be applied to the known mechanism of DSB formation / and the relationship with chromosome axis.

While examples of positive correlation of interference strength and genome-wide recombination rate are less common; the within sex comparison of two breeds of cattle with different genome wide recombination rates  [@ma2015_cattle], between lab and wild mice of Peromyscus leucopus, [@peterson2019].

we propose that --- this is due to difference in scale (resolution)
and how these two metrics are measured at the genome wide vs single chromosome (bivalent) level.

<explain this difference>

<**the models**>

@@

 We propose that the **spindle based selection model** support the evolution of stronger interference strength via selection onthe amount of sister cohesion connecting homologs (figure).  -- and this can evolve with increased crossover number

<we don't know the chromosome identities, but based on the propotions of chromosome classes this is the logical conclusion.

The evolution of gwRR in our study is driven by a transition of recombination landscapes in a propotion of chromosomes, from single crossovers to two crossovers --

All else being equal, axis length) -- interference strength would be expected to be equal or even weaker

Our model suggests that an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs and could change the chromsome structure of these bivalents when they are aligned and bi-oriented. This model gives rise to the possibility that the chromosome structure at metaphase is the target of selection and changing the recombination landscape is an outcome  pleiotropic. This interpertation has potential to fit with hypotheses mentioned above for explaining other results regarding the evolution of the recombination landscape.

We have focused on the model involving direct selection on the recombination pathway, that is the impact on the meotic process on an indiviual, but we make note on models centering on indirect selection or fitness outcomes of offspring which could also explain our results. Models from Goldstein et al, reviewed in [@ottoPaysuer2019] and Veller 2017?, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of their model is that the number of crossovers per chromosome (genetic unit) is kept constant. The range of crossovers per chromosome is quite small, 1-3, for the majority of chromosomes [@ottoPaysuer2019, @stapley_variation_2017]. Constraining the number of crossovers per chromosome in models may fit empiracal data better than those where recombiantion rate across an abstracted genetic space is unconstrained.

>>>>

We performed hypothesis testing of models from the literature and adapted here fit the main results from this study. We choose to examine three modifier models and two functional and cell physiology models (table X). It is a bit challenging to test all of these models together since they were built to describe different patterns of recombination variation. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the COM model was built to describe the interference pattern and difference in the recombination landscape between the sexes. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

The direction of heterochiasmy under the two locus modifier model is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) that the driver acts (see Table 1 of @brandvain2012scrambling).

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle before starting the irreversible metaphase-to-anaphase transition (cite).  A stable chromosome structure, known as bi-orientation, requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion (cite). Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, Subrahman). 
These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite). This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

@@

Our results of greater between cell variation in females are supported by empiracal results from human and mouse [@lynn2002, @gruhn2013]. <More details on the empirarcal pattern>

While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers (Schuh Ellenburg 2007). This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite).

<stabilization -- bi-oriented, robust SAC response, SAC acts to correctly time metaphase to anaphase transition -- only after correct alignment and bi-orientation of bivalents in the metaphase cell, -- spindle dynamics much different due to asymmetrical division Mogessie et al 2018, Mogessie Schuh --actin and MT are apart of oocyte spindle>

<Bennabi et al 2016 --spindle assembly inoocytes, sun and kim 2011 -- SAC and regulators, Kolano et al 2012 --error prone oocytes, Manil-Segalen -- forces through MT breaking chromosomes, Mogessie et al 2018 -- meiotic spindle (big)review>

@@

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE model predicts sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The two mechanical models both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (cite). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate (cite). Different timing of cell cycle between oogenesis and spermatogenesis may impose the the different selective pressures on sister cohesion. Faster spermatogenesis may select for synchronization of the separation homologs. While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle.

@@

Our results from musculus^MSM^ and musculus^PWD^ show that the SC length and chromatin compaction is uncoupled from the direction of heterochiasmy. These results slightly depart from predictions which nominate chromatin organization (axis length) as the primary driver of heterochiasmy [@petkov2007].  While the relationship may hold for more general heterochiasmy patterns across broad scale comparisons, these results indicate that rapid male specific  evolution in gwRR, is through a different trait than chromatin compaction.

<The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002, @gruhn2013], But there's generally a lack of data for other systems see [@cahoonLibuda2019]).>
Only the COM model predicts sexual dimorphism in chromatin structure will be longer in oocytes due to greater cell volume. This prediction model might fit broader pattern such as in *Arabidopsis* where pollen is the larger cell and has longer axis length [@cahoonLibuda2019].

@@

Examples of evolution of interference strength are rare and may be complicated by changes in the underlying karyotype structure (Segura et al 2014). The general pastern of empirical measures of interference strength is that it has a negative correlation to the genome wide recombination rate [@ottoPaysuer2019]. This fits the logic of the chromosome axis acting as the 'real estate' for DSBs and aligns with known molecular mechanisms (cite). Two examples of the opposite direction, a positive correlation of interference strength and genome-wide recombination rate, involve descriptions of observations at the single bivalent level; the between lab-raised and wild mice of *Peromyscus leucopus* from [@peterson2019] or large effects on genome wide recombination rates across a short evolutionary differences; the within-sex comparison of two breeds of cattle with different genome wide recombination rates [@ma2015_cattle]. We propose that there is a difference in resolution and power between these two groups of empirical results with the positive and negative relationship between genome wide recombination rates and interference strength.

We propose that the **spindle based selection model** can explain the positive correlation and evolution of stronger interference strength via selection on the amount of sister cohesion connecting homologs (figure X, table X). The evolution of genome wide recombination rates in our study is driven by a transition of recombination landscapes from single crossovers to two crossovers. All else being equal, interference strength would be expected to be equal or even weaker. Our results show that the two crossovers are spaced further apart when SC length is controlled for and chromosome size effects are minimized, in the strain with more two crossover chromosomes. Assuming that chromatin compaction between pachtyene and metaphase is uniform along chromosomes, an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs (Figure X). This model gives rise to the possibility that the chromosome structure at metaphase is the target of selection  which is achieved by changing the average recombination landscape across bivalents within the cell. This interpretation has potential to fit with hypotheses mentioned above for explaining other results regarding the evolution of the recombination landscape.

@@

We have focused on the model involving direct selection on the recombination pathway, that is the impact on the meiotic process on an individual, but we make note on models centering on indirect selection or fitness outcomes of offspring which could also explain our results. Models from Goldstein et al, reviewed in [@ottoPaysuer2019] and Veller 2017, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of their model is that the number of crossovers per chromosome is kept constant. The range of crossovers per chromosome is quite small, 1-3, for the majority of chromosomes [@ottoPaysuer2019, @stapley_variation_2017]. Constraining the number of crossovers per chromosome in models may fit empirical data better than those where recombination rate across an abstracted genetic space is unconstrained.

<Still working on Future Steps>
